Spatial Reasoning and Statistical Graphics

### 2014-11-10 Paper/Draft1/TestingVisualAptitude.Rnw
lineup types
Working memory and cognitive load explanation

\todo[inline]{Intro about statistical graphics, lineups, ...}
\subsection{Spatial Reasoning and Statistical Graphics}
Statistical graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs:  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed ~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior\cite{zhang1997nature}. This constraint reduces memory load and makes more cognitive resources available for other tasks, but also implicates visual ability as an important factor in graph comprehension.

@@

<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\subsection{Lineup Types}
\newtext{
We can also compare participants' performance on specific types of lineup plots compared with their scores on the visual aptitude tests, for instance, accuracy on lineups which require mental rotation may be related to performance on the card rotation task. Examples lineups for each type of plot are provided in Appendix \ref{app:lineuptypes}.
}

>>>>

\todo[inline]{Intro about statistical graphics, lineups, ...}
\subsection{Spatial Reasoning and Statistical Graphics}
Statistical graphics provide quick summaries of data, models, and results, but these displays may not be equally useful to all viewers. Mathematical ability is important, even for the simplest graphs:  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. Spatial ability becomes more important, however, when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed ~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior\cite{zhang1997nature}. 
\newtext{Graphics utilize higher-bandwidth visual pathways to encode information\cite{baddeley1974working}, allowing viewers to quickly and intuitively encode multiple dimensions of numerical relationships; however, this higher-bandwidth approach also depends on the viewer's visual and spatial reasoning ability. Well-designed graphics present the most important features of the data while minimizing less-important features, guiding the viewer towards conclusions that are meaningful in context and supported by the data while maximizing the information encoded in working memory. Under this framework, well-designed graphics reduce memory load and make more cognitive resources available for other tasks (such as drawing conclusions from the data), at the cost of depending on certain visuospatial reasoning abilities. }

@@

<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\subsection{Lineup Types}
\newtext{The survey consisted of three sets of different types of lines. Examples lineups for each type of plot are provided in Appendix \ref{app:lineuptypes}. 
}
\newtext{
We can also compare participants' performance on specific types of lineup plots compared with their scores on the visual aptitude tests, for instance, accuracy on lineups which require mental rotation may be related to performance on the card rotation task. 
}
